An exposition on the propriety of restricted Boltzmann machines

### 2016-09-11 writing/draft.Rmd
first stab at instability edits

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes. An exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distibrution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$. Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.

@@

@schweinberger2011instability introduces a related concept, *instability*, which means [**TODO: that small changes in the natural parameters result in large changes in probability masses**]. This is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. @schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model where [**TODO: notation goes here**].

\begin{defn}
A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
\end{align}

Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
\end{align}
\end{defn}

For RBMs, an alaogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.

@@

Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.

\begin{lemma}[Schweinberger (2011)]
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
\end{lemma}

Thus, if a discrete exponential family model is unstable, then it is also degenerate.

>>>>

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes. An exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$. Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.

@@

@schweinberger2011instability introduces a related concept, *instability*. Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. Consider discrete exponential families of distributions with probability mass functions of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where, $\eta(\boldsymbol \theta)$ is a vector of natural parameters, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of sufficient statistics based on $\boldsymbol x \in \mathcal{X}^N$ for $\mathcal{X}$ a discrete set of $M$ elements and $N$ degrees of freedom. In the random graph application, $\mathcal{X}^N$ may correspond to the number of relationships among $n$ nodes in a graph. In the instance of RBMs, $N$ refers to the number of hiddens and visible nodes included in the model, $N = H + V$. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}^N} \exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right))$.

@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.

\begin{defn}
A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] > CN \text{ for all } N > N_C \label{eq:unstable}
\end{align}

Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] \le CN \text{ for all } N > N_C. \label{eq:stable}
\end{align}
\end{defn}

For RBMs, an analagous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.

@@

Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.

\begin{lemma}[Schweinberger (2011)]
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x)\right) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)]$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
\end{lemma}

Thus, if a discrete exponential family model is unstable, then it is also near-degenerate. Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
